Linking watershed nutrient loading to estuary water quality with Generalized Additive Models

### 2023-08-11 Schramm-2023-06-PeerJ/Schramm-2023-08-PeerJ.tex
resubmission ready

\begin{abstract}
Evaluating estuary water quality responses to reductions (or increases)
in nutrient loading attributed to on the ground management actions can
be challenging due to the strong influence of environmental drivers on
nutrient loads and non-linear relationships. This study applied
semiparametric approaches to calculate watershed nutrient loads and
assess responses in estuary water quality to seasonally-adjusted
freshwater inflow and flow-adjusted nutrient loads in Lavaca Bay, Texas.
Lavaca Bay is a secondary embayment on the Texas coast displaying early
potential for eutrophication and water quality degradation. Use of
flow-adjusted nutrient loads allowed the study to evaluate the response
in water quality to changes in nutrient loads driven by anthropogenic
sources. Cross-validation indicated that, despite data constraints,
semiparametric models performed well at nutrient load prediction. Based
on these models, delivered annual nutrient loads varied substantially
from year to year. In contrast, minimal changes in flow-normalized loads
indicate that nutrient loadings were driven by natural variation in
precipitation and runoff as opposed to changes in management of nonpoint
sources. Models indicated no evidence of long-term changes in dissolved
oxygen or chlorophyll-\emph{a} within Lavaca Bay. However, site specific
long-term increases in both organic and inorganic nitrogen are
concerning for their potential to fuel eutrophication. Further analysis
found freshwater inflow had strong influences on nutrient and
chlorophyll-\emph{a} concentrations but there was no evidence that
changes in watershed nutrient loading explained additional variation in
dissolved oxygen or chlorophyll-\emph{a} concentrations. In addition to
providing a baseline assessment of watershed nutrient loading and water
quality responses in the Lavaca Bay watershed, this study provides
methodological support for the use of semiparametric models in load
regression models and estuary assessments.
% Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text.
\end{abstract}

@@

The goal of this project is to assess watershed nutrient loading and the
resulting water quality responses in Lavaca Bay, Texas. Lavaca Bay is a
secondary bay in the larger Matagorda Bay system located roughly halfway
between Houston, Texas and Corpus Christi, Texas. Lavaca Bay faces
substantial challenges associated with legacy contamination but general
water quality parameters such as dissolved oxygen (DO), nutrients, and
biological parameters have been well within state water quality
standards. Despite largely meeting state designated water quality
thresholds, there have been concerning declines in abundance, biomass,
and diversity of benthic fauna in Lavaca Bay
\autocite{beserespollackLongtermTrendsResponse2011}. These declines are
partially attributed to reductions in freshwater inflow and changes in
estuary salinity and are indicative of an already stressed system
\autocite{beserespollackLongtermTrendsResponse2011,palmerImpactsDroughtsLow2015,montagnaAssessmentRelationshipFreshwater2020}.
More recently, significant linear increases in total phosphorus (TP),
orthophosphate, total Kjeldahl nitrogen (TKN), and chlorophyll-\emph{a}
concentrations were identified at monitoring sites within Lavaca Bay
\autocite{bugica_water_2020}. These trends in nutrient concentrations
are concerning due to the role of nitrogen as a limiting factor for
primary production in many Texas estuaries
\autocite{gardnerNitrogenFixationDissimilatory2006,houTransformationFateNitrate2012,doradoUnderstandingInteractionsFreshwater2015,paudelRelationshipSuspendedSolids2019,wetz_exceptionally_2017}
and the ramifications that changes in nitrogen loadings could have for
productivity and eutrophication in Lavaca Bay.

@@

Water quality data for the two freshwater and three estuary locations
were obtained from the Texas Commission on Environmental Quality (TCEQ)
Surface Water Quality Monitoring Information System. Data submitted
through the system are required to be collected under Quality Assurance
Project Plans and lab method procedures outlined by the TCEQ's
procedures manual. These operating procedures ensure consistent
collection and laboratory methods are applied between samples collected
by different entities and under different projects. All sites had
varying time periods and availability of water quality data. For
freshwater locations, TP from January 2000 through December 2020 and
nitrate-nitrogen (NO\textsubscript{3}) data from January 2005 through
December 2020 were downloaded (Table~\ref{tab:table1}). Less than
5-years of total nitrogen and TKN concentration data were available at
the freshwater sites and deemed insufficient to develop load estimation
models
\autocite{horowitzEvaluationSedimentRating2003,snelderEstimationCatchmentNutrient2017}.
The three estuary sites included an upper-bay site near the outlet of
the Lavaca River system (TCEQ-13563), a mid-bay site (TCEQ-13383), and
the lower-bay site near the mouth of the Bay (TCEQ-13384). For estuary
locations, we obtained data for TP, Nitrite+Nitrate
(NO\textsubscript{\emph{X}}), TKN, chlorophyll-\emph{a}, and DO
concentrations from January 2005 through December 2020
(Table~\ref{tab:table2}). All water quality data was collected
approximately quarterly.

@@

Left-censored data were not uncommon in this dataset. Several methods
are available to account for censored data. We transformed left-censored
nutrient concentrations to one-half the detection limit. Although this
simple approach can introduce bias
\autocite{hornungEstimationAverageConcentration1990}, we considered it
acceptable because high concentrations and loadings are associated with
high-flow events and low-flow/low-concentration events will account for
a small proportion of total loadings
\autocite{mcdowell_implications_2021}.

@@

\caption{Aggregated estimated annual and flow-normalized annual NO\textsubscript{3} and TP loads for USGS-08164000 and USGS-08164525.}\label{fig:fig3}
\end{figure}

@@

There was no evidence of long-term changes in TP or DO concentrations at
any Lavaca Bay site (Fig.~\ref{fig:fig5}). The upper-bay site,
TCEQ-13563, had evidence of a long-term linear increase in
NO\textsubscript{\emph{X}} while chlorophyll-\emph{a} decreased from
2005 through 2014 (Fig.~\ref{fig:fig5}). NO\textsubscript{\emph{X}}
concentration at the mid-bay site, TCEQ-13383, displayed a periodic
pattern that is indicative of a strong influence from inflow or
precipitation. The temporal GAMs did not provide evidence of long-term
trends in any of the water quality constituents at the lower-bay
TCEQ-13384 site.

@@

Freshwater inflow provided additional explanation for changes in TP and
NO\textsubscript{\emph{X}} concentration at all of the Lavaca Bay sites
according to AIC\textsubscript{c} and model probability values
(Table~\ref{tab:table4}). TCEQ-13563, the site closest to the river
outlet, was the only site that had improvements in the explanations of
DO and TKN concentration with the inclusion of inflow. Both TCEQ-13563
and TCEQ-13383, the mid-bay site, saw improvements in explanations for
variations in chlorophyll-\emph{a} with the inclusion of freshwater
inflow. The addition of nutrient loads (both TP and NO\textsubscript{3})
terms did not provide additional explanation for changes in
chlorophyll-\emph{a} or DO concentrations. Inclusion of TP loads
provided additional explanation of TP concentrations at the upper- and
mid-bay sites, TCEQ-13563 and TCEQ-13383. Inclusion of
NO\textsubscript{3} loads only provided marginal improvements in the
explanation of NO\textsubscript{\emph{X}} concentration at the lower-bay
TCEQ-13384 site.

@@

Increases in freshwater inflow resulted in nearly linear increases in TP
and NO\textsubscript{\emph{X}} concentration at all three sites
(Fig.~\ref{fig:fig6}). At the upper-bay TCEQ-13563 site, increases in
freshwater inflow initially increased chlorophyll-\emph{a} and DO
concentration but concentrations leveled and potentially decreased at
higher flows. The mid-bay TCEQ-13383 site had nearly linear increases in
chlorophyll-\emph{a} concentration in response to increased freshwater
inflow. There was no evidence of freshwater inflow effects on
chlorophyll-\emph{a}, TKN, or DO at the lower-bay TCEQ-13384 site.

@@

Increased TP loads resulted in nearly linear increases of TP
concentration at the upper- and mid-bay sites, TCEQ-13563 and TCEQ-13383
respectively (Fig.~\ref{fig:fig7}). The relative effect size appeared
much smaller than the effect of freshwater inflow alone. Increased
NO\textsubscript{3} loads only showed an effect at the lower-bay
TCEQ-13384 site. The effect was quite small compared to streamflow and
provided only small improvements to the model (Table~\ref{tab:table4}).
As noted above, nutrient loadings did not provide any explanation in
changes in the remaining assessed water quality parameters.

@@

\caption{Fitted splines from the nutrient loading GAMs display the marginal smoothed effect of TP and NO\textsubscript{3} loads on TP (A) and NO\textsubscript{\textit{X}} (B) concentrations at each site in Lavaca Bay.}\label{fig:fig7}
\end{figure}

@@

The strong positive effect of freshwater inflow on
NO\textsubscript{\emph{X}}, TKN, and TP concentration are suggestive of
nonpoint watershed sources, consistent with watershed uses and with
other studies relating freshwater inflow with nutrient concentrations in
Lavaca Bay and other estuaries
\autocite{russell_effect_2006,caffreyHighNutrientPulses2007,peierlsNonmonotonicResponsesPhytoplankton2012,palmerImpactsDroughtsLow2015,ciraPhytoplanktonDynamicsLowinflow2021}.
Inflow had a non-linear relationship with TKN at the two upstream sites,
with TKN increasing as freshwater inflow transitioned from low to
moderate levels. At higher freshwater inflows, the effect was
attenuated, possibly indicating a flushing effect at higher freshwater
inflow. No relationship between TKN and freshwater inflow was observed
at TCEQ-13384 located in the lower reach of Lavaca Bay. The results
suggest that tidal flushing might dilute TKN and act as a control on the
effects of freshwater inflow in the lower reaches of Lavaca Bay. The
results are also consistent with previous work that suggest processing
of organic loads in the upper Lavaca Bay or tidal portions of the Lavaca
River reduce transport of nutrients to the lower reaches of Lavaca Bay
\autocite{russell_effect_2006}.

Freshwater inflow also displayed a strong positive effect on
chlorophyll-\emph{a} at the upper- and mid-bay sites. The upper-bay
site, TCEQ-13563, showed decreases in chlorophyll-\emph{a} at the
highest freshwater inflow volumes. Freshwater flushing or increases in
turbidity are associated with decreases in chlorophyll-\emph{a} in other
estuaries
\autocite{peierlsNonmonotonicResponsesPhytoplankton2012,cloernPhytoplanktonPrimaryProduction2014}.
No relationships between inorganic nitrogen or TP loadings with
chlorophyll-\emph{a} were observed. Due to the lack of TKN loading
information, no assessment between organic nitrogen loads and
chlorophyll-\emph{a} were possible.

@@

Overall, this study suggests that DO and chlorophyll-\emph{a}
concentrations have been relatively stable in Lavaca Bay. Site-specific
increases in TKN and NO\textsubscript{\emph{X}} concentrations are a
cause of concern for increasing risks of eutrophication within Lavaca
Bay which might be currently attenuated by changes in freshwater flow,
turbidity, and other physical processes. While loading models indicate
that there are large annual fluctuations in NO\textsubscript{3} loads,
these changes have been largely driven environmental conditions (changes
in runoff and river discharge). These models also provide evidence that
estuary NO\textsubscript{\emph{X}} and TP concentrations are strongly
driven by freshwater inflow and to a lesser extent fluctuation in
flow-adjusted riverine loadings. Site-specific changes in the
relationships between freshwater inflow and responses in both
chlorophyll-\emph{a} and TKN concentrations are indicative nutrient
processing and or tidal flushing effects moving from the river discharge
point to the mouth of Lavaca Bay. This study does not completely explain
site specific increases in NO\textsubscript{\emph{X}} and TKN
concentrations in Lavaca Bay. The freshwater study sites did not
quantify nutrient loadings from tidal contribution areas or ungauaged
watersheds. Nutrient contributions from wastewater facility discharges,
septic systems, and stormwater could be considerable contributors to
nutrient loadings in Lavaca Bay since they are not processed by a tidal
river reach prior to entering the Bay. The Garcitas-Arenosa Creek,
Placedo Creek, and Cox Bay subwatersheds are currently undersampled but
compose approximately 27\% of the watershed area. The contribution of
nutrient loadings from these undersampled areas is unknown.

\hypertarget{limitations}{%
\subsection*{Limitations}\label{limitations}}
\addcontentsline{toc}{subsection}{Limitations}

@@

This study, consistent with others along the Texas coast, found strong
effects of freshwater flow on nutrient and chlorophyll-\emph{a}
concentrations. DO concentrations, dominated by seasonal effects, did
not show strong direct responses to freshwater flow. Small variances in
flow-adjusted nutrient loads indicate that (1) there have been limited
changes in non-point sources of nutrients and (2) that there is not
strong evidence that those small changes have had effects on
chlorophyll-\emph{a} or dissolved oxygen in Lavaca Bay. Although this
study did not identify changes in DO or chlorophyll-\emph{a}
concentrations in Lavaca Bay, site specific increases in
NO\textsubscript{\emph{X}} and TKN are a cause for water quality
concern. The study provides a baseline assessment for future water
quality management activities in the watershed. In order to effectively
track and link improvements or degradation of water quality conditions
in Lavaca Bay and other coastal Texas watersheds with on the ground
efforts, more robust sampling networks are needed to improve spatial
coverage of undersampled areas and explicitly incorporate flow-biased
sampling.

>>>>

\begin{abstract}
Evaluating estuary water quality responses to reductions (or increases)
in nutrient loading attributed to on the ground management actions can
be challenging due to the strong influence of environmental drivers on
nutrient loads and non-linear relationships. This study applied
generalized additive models to calculate watershed nutrient loads and
assess responses in estuary water quality to seasonally-adjusted
freshwater inflow and flow-adjusted nutrient loads in Lavaca Bay, Texas.
Lavaca Bay is a secondary embayment on the Texas coast displaying early
potential for eutrophication and water quality degradation. Use of
flow-adjusted nutrient loads allowed the study to evaluate the response
in water quality to changes in nutrient loads driven by anthropogenic
sources. Cross-validation indicated that, despite data constraints,
semiparametric models performed well at nutrient load prediction. Based
on these models, delivered annual nutrient loads varied substantially
from year to year. In contrast, minimal changes in flow-normalized loads
indicate that nutrient loadings were driven by natural variation in
precipitation and runoff as opposed to changes in management of nonpoint
sources. Models indicated no evidence of long-term changes in dissolved
oxygen or chlorophyll-\emph{a} within Lavaca Bay. However, site specific
long-term increases in both organic and inorganic nitrogen are
concerning for their potential to fuel eutrophication. Further analysis
found freshwater inflow had strong influences on nutrient and
chlorophyll-\emph{a} concentrations but there was no evidence that
changes in watershed nutrient loading explained additional variation in
dissolved oxygen and limited evidence that watershed nutrient loadings
explained chlorophyll-\emph{a} concentrations. In addition to providing
a baseline assessment of watershed nutrient loading and water quality
responses in the Lavaca Bay watershed, this study provides
methodological support for the use of semiparametric models in load
regression models and estuary assessments.
% Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text.
\end{abstract}

@@

The goal of this project is to assess watershed nutrient loading and the
resulting water quality responses in Lavaca Bay, Texas. Lavaca Bay is a
secondary bay in the larger Matagorda Bay system located roughly halfway
between Houston, Texas and Corpus Christi, Texas. Lavaca Bay faces
substantial challenges associated with legacy contamination but general
water quality parameters such as dissolved oxygen (DO), nutrients, and
biological parameters have been well within state water quality
standards. Despite largely meeting state designated water quality
thresholds, there have been concerning declines in abundance, biomass,
and diversity of benthic fauna in Lavaca Bay
\autocite{beserespollackLongtermTrendsResponse2011}. These declines are
partially attributed to reductions in freshwater inflow and changes in
estuary salinity and are indicative of an already stressed system
\autocite{beserespollackLongtermTrendsResponse2011,palmerImpactsDroughtsLow2015,montagnaAssessmentRelationshipFreshwater2020}.
More recently, significant linear increases in total phosphorus (TP),
orthophosphate, total Kjeldahl nitrogen (TKN), and chlorophyll-\emph{a}
concentrations were identified at monitoring sites within Lavaca Bay
\autocite{bugica_water_2020}. While significant nitrogen increases were
only identified in TKN, both organic and inorganic forms of nitrogen can
serve as potential indicators of eutrophication status and estuary
health \autocite{jessenMarineEutrophication2015}. Nitrogen is generally
considered the limiting factor for primary production in many Texas
estuaries
\autocite{gardnerNitrogenFixationDissimilatory2006,houTransformationFateNitrate2012,doradoUnderstandingInteractionsFreshwater2015,paudelRelationshipSuspendedSolids2019,wetz_exceptionally_2017}.
However, estuaries can display nitrogen and phosphorus co-limitation
that may vary temporally and spatially
\autocite{elserGlobalAnalysisNitrogen2007,conleyControllingEutrophicationNitrogen2009},
emphasizing a need to assess a variety of nutrient species.

@@

Water quality data for the two freshwater and three estuary locations
were obtained from the Texas Commission on Environmental Quality (TCEQ)
Surface Water Quality Monitoring Information System. Data submitted
through the system are required to be collected under Quality Assurance
Project Plans and lab method procedures outlined by the TCEQ's
procedures manual. These operating procedures ensure consistent
collection and laboratory methods are applied between samples collected
by different entities and under different projects. All sites had
varying time periods and availability of water quality data. For
freshwater locations, TP from January 2000 through December 2020 and
nitrate-nitrogen (NO\textsubscript{3}) data from January 2005 through
December 2020 were downloaded (Table~\ref{tab:table1}). Less than
5-years of total nitrogen and TKN concentration data were available at
the freshwater sites and deemed insufficient to develop load estimation
models
\autocite{horowitzEvaluationSedimentRating2003,snelderEstimationCatchmentNutrient2017}.
The three estuary sites included an upper-bay site near the outlet of
the Lavaca River system (TCEQ-13563), a mid-bay site (TCEQ-13383), and
the lower-bay site near the mouth of the Bay (TCEQ-13384). For estuary
locations, we obtained data for TP, Nitrite+Nitrate
(NO\textsubscript{\emph{X}}), TKN, chlorophyll-\emph{a}, and DO
concentrations from January 2005 through December 2020
(Table~\ref{tab:table2}). Freshwater nutrient samples were collect 0.3
meters below the surface or halfway between the surface and bottom if
depth was insufficient. Estuary nutrient samples were all collected 0.3
meters below the surface. Estuary DO measurements were made at 3
locations along the water column (0.3 meters, mid-depth, and bottom) and
averaged. All water quality data was collected approximately quarterly.

@@

Left-censored data were not uncommon in this dataset (4\% of TP samples
and 20\% of NO\textsubscript{3} samples). Several methods are available
to account for censored data. We transformed left-censored nutrient
concentrations to one-half the detection limit. Although this simple
approach can introduce bias
\autocite{hornungEstimationAverageConcentration1990}, we considered it
acceptable because high concentrations and loadings are associated with
high-flow events and low-flow/low-concentration events will account for
a small proportion of total loadings
\autocite{mcdowell_implications_2021}.

@@

\caption{Aggregated estimated annual and flow-normalized annual NO\textsubscript{3} and TP loads for the Lavaca (USGS-08164000) and and Navidad (USGS-08164525) Rivers.}\label{fig:fig3}
\end{figure}

@@

There was no evidence of long-term changes in TP or DO concentrations at
any Lavaca Bay site (Fig.~\ref{fig:fig5}). The upper-bay site
(TCEQ-13563) had evidence of a long-term linear increase in
NO\textsubscript{\emph{X}} while chlorophyll-\emph{a} decreased from
2005 through 2014 (Fig.~\ref{fig:fig5}). NO\textsubscript{\emph{X}}
concentration at the mid-bay site (TCEQ-13383) displayed a periodic
pattern that is indicative of a strong influence from inflow or
precipitation. The temporal GAMs did not provide evidence of long-term
trends in any of the water quality constituents at the lower-bay site
(TCEQ-13384).

@@

Freshwater inflow provided additional explanation for changes in
NO\textsubscript{x} and chlorophyll-\emph{a} at all of the Lavaca Bay
sites according to AIC\textsubscript{c} and model probability values
(Table~\ref{tab:table4}). Freshwater inflow also explained additional
variation in TP at the upper- and mid-bay sites but not the lower-bay
site. Freshwater inflow did not explain additional variation in TKN
concentrations at any of the sites. The upper- and lower-bay sites saw
improvements in explanation of DO concentration with the inclusion of
inflow.

Inclusion of TP loads provided additional explanation of TP
concentrations at the upper- and mid-bay sites. Inclusion of
NO\textsubscript{3} loads did not provide model explanatory improvements
at any of the sites. The addition of nutrient loads (both TP and
NO\textsubscript{3}) terms did not provide additional explanation for
changes in DO concentrations but did provide model improvement for the
chlorophyll-\emph{a} model at the mid-bay site.

Increases in aggregated freshwater inflow resulted in increases in TP at
the upper- and mid-bay sites and increases in NO\textsubscript{\emph{X}}
and chlorophyll-\emph{a}concentration at all three sites
(Fig.~\ref{fig:fig6}). Linear increases in DO concentration were
observed with increasing flow at the upper- and lower-bay sites. TKN
showed no response to changes in inflow at any of the sites.

Increased TP loads resulted in nearly linear increases of TP
concentration at the upper- and mid-bay sites (Fig.~\ref{fig:fig7}). The
relative effect size appeared much smaller than the effect of freshwater
inflow alone. Increased NO\textsubscript{3} loads did effect
NO\textsubscript{x} concentrations at any site. Linear increases in
chlorophyll-\emph{a} were observed in response to increased TP loads at
the mid-bay site, but not the other sites or in response to
NO\textsubscript{3} loads.

@@

@@

@@

\caption{Fitted splines from the nutrient loading GAMs display the marginal smoothed effect of 20-day aggregated TP and NO\textsubscript{3} flow-normalized loads on (A) TP, (B) NO\textsubscript{\textit{X}}, and (C,D) chlorophyll-\textit{a} concentrations at each site in Lavaca Bay.}\label{fig:fig7}
\end{figure}

@@

The strong positive effect of freshwater inflow on
NO\textsubscript{\emph{X}} and TP concentration are suggestive of
nonpoint watershed sources, consistent with watershed uses and with
other studies relating freshwater inflow with nutrient concentrations in
Lavaca Bay and other estuaries
\autocite{russell_effect_2006,caffreyHighNutrientPulses2007,peierlsNonmonotonicResponsesPhytoplankton2012,palmerImpactsDroughtsLow2015,ciraPhytoplanktonDynamicsLowinflow2021}.
Inflow had a non-linear relationship with NO\textsubscript{x}, with
NO\textsubscript{x} increasing as freshwater inflow transitioned from
low to moderate levels. At higher freshwater inflows, the
NO\textsubscript{x} decreased at the mid- and lower-bay sites, possibly
indicating a flushing effect at higher freshwater inflow. No
relationship between TKN and freshwater inflow was observed at any of
the sites. The results coincide with previous work that suggest
processing of organic loads in the tidal portions of the Lavaca River
reduce transport of nutrients to the lower reaches of Lavaca Bay
\autocite{russell_effect_2006}.

Freshwater inflow also displayed positive effects on
chlorophyll-\emph{a} at each site, with the largest effect at the
lower-bay site. The lower-bay site also showed an attenuation in in
chlorophyll-\emph{a} at the highest freshwater inflow volumes.
Freshwater flushing or increases in turbidity are associated with
limitations or decreases in chlorophyll-\emph{a} in other estuaries
\autocite{peierlsNonmonotonicResponsesPhytoplankton2012,cloernPhytoplanktonPrimaryProduction2014}.
No relationships between inorganic nitrogen and chlorophyll-\emph{a}
were observed. However, a small positive effect between flow-normalized
TP loads and chlorophyll-\emph{a} concentrations were detected at the
mid-bay site. Although, this is not suggestive of a phosphorus
limitation in the Bay, it is supportive of prior work emphasizing the
importance of controlling both nitrogen and phosphorus runoff to protect
water quality \autocite{conleyControllingEutrophicationNitrogen2009}.
Due to the lack of TKN loading information, no assessment between
organic nitrogen loads and chlorophyll-\emph{a} were possible.

@@

Overall, this study suggests that DO and chlorophyll-\emph{a}
concentrations have been relatively stable in Lavaca Bay. Site-specific
increases in TKN and NO\textsubscript{\emph{X}} concentrations are a
cause of concern for increasing risks of eutrophication within Lavaca
Bay which might be currently attenuated by changes in freshwater flow,
turbidity, and other physical processes. While loading models indicate
that there are large annual fluctuations in NO\textsubscript{3} loads,
these changes have been largely driven environmental conditions (changes
in runoff and river discharge). These models also provide evidence that
estuary NO\textsubscript{\emph{X}} and TP concentrations are strongly
driven by freshwater inflow and to a lesser extent fluctuation in
flow-normalized riverine loadings. Site-specific changes in the
relationships between freshwater inflow and responses in both
chlorophyll-\emph{a} and NO\textsubscript{x} concentrations are
indicative of nutrient processing and or tidal flushing effects moving
from the river discharge point to the mouth of Lavaca Bay. This study
does not completely explain site specific increases in
NO\textsubscript{\emph{X}} and TKN concentrations in Lavaca Bay. The
freshwater study sites did not quantify nutrient loadings from tidal
contribution areas or ungauaged watersheds. Nutrient contributions from
wastewater facility discharges, septic systems, and stormwater could be
considerable contributors to nutrient loadings in Lavaca Bay since they
are not processed by a tidal river reach prior to entering the Bay. The
Garcitas-Arenosa Creek, Placedo Creek, and Cox Bay subwatersheds are
currently undersampled but compose approximately 27\% of the watershed
area. The contribution of nutrient loadings from these undersampled
areas is unknown.

\hypertarget{future-management-and-research-needs}{%
\subsection*{Future Management and Research
Needs}\label{future-management-and-research-needs}}
\addcontentsline{toc}{subsection}{Future Management and Research Needs}

@@

This study, consistent with others along the Texas coast, found strong
effects of freshwater flow on nutrient and chlorophyll-\emph{a}
concentrations. DO concentrations, dominated by seasonal effects, did
not show strong direct responses to freshwater flow. Small variances in
flow-adjusted nutrient loads indicate that (1) there have been limited
changes in non-point sources of nutrients and (2) that there is not
strong evidence that those small changes have had extensive effects on
chlorophyll-\emph{a} or dissolved oxygen in Lavaca Bay. Although this
study did not identify changes in DO or chlorophyll-\emph{a}
concentrations in Lavaca Bay, site specific increases in
NO\textsubscript{\emph{X}} and TKN are a cause for water quality
concern. The study provides a baseline assessment for future water
quality management activities in the watershed. In order to effectively
track and link improvements or degradation of water quality conditions
in Lavaca Bay and other coastal Texas watersheds with on the ground
efforts, more robust sampling networks are needed to improve spatial
coverage of undersampled areas and explicitly incorporate flow-biased
sampling.
